---
title: "calibration"
format: revealjs
---

# Calibration **⚖️**

-   How well the model is "calibrated": Patients with a probability of about 0.2 are expected to have a proportion of about 0.2 observed events.

-   In order to asses calibration we need to use quantiles of estimated probabilities (discrete version of calibration) or some kind of smoothing algorithm.

-   The main idea is to visually inspect for similarity with the linear line of 45 degrees.

-   Visual inspection might be problematic, but from our experience it is a good-enough practice.

## Calibration **⚖️**

### Why should we care?

-   An accurate model in terms of discrimination might produce uncalibrated estimated probabilities what will lead to poor decisions.

-   Logistic Regression is calibrated by default, but if you use fancy ML prediction models they might not be calibrated.

## Calibration **⚖️**

### What if the model is not calibrated?

-   Uncalibrated models can be fixed by remodeling the predictions with a simple logistic regression (recalibration).

-   Python users might use [`sklearn.calibration.CalibratedClassifierCV`](https://scikit-learn.org/stable/modules/generated/sklearn.calibration.CalibratedClassifierCV.html).

-   Optimizing the logloss function will produce calibrated model.

## **Discrete** Calibration **⚖️**

<table style="width:98%;">
<colgroup>
<col style="width: 8%" />
<col style="width: 8%" />
<col style="width: 8%" />
<col style="width: 8%" />
<col style="width: 8%" />
<col style="width: 8%" />
<col style="width: 8%" />
<col style="width: 8%" />
<col style="width: 8%" />
<col style="width: 8%" />
<col style="width: 8%" />
</colgroup>
<tbody>
<tr>
<td style="text-align: center;"><p><strong>p̂</strong></p></td>
<td style="text-align: center;"><p>0.11</p></td>
<td style="text-align: center;"><p>0.15</p></td>
<td style="text-align: center;"><p>0.18</p></td>
<td style="text-align: center;"><p>0.29</p></td>
<td style="text-align: center;"><p>0.31</p></td>
<td style="text-align: center;"><p>0.33</p></td>
<td style="text-align: center;"><p>0.45</p></td>
<td style="text-align: center;"><p>0.47</p></td>
<td style="text-align: center;"><p>0.63</p></td>
<td style="text-align: center;"><p>0.72</p></td>
</tr>
<tr>
<td style="text-align: center;"><p><strong>R</strong></p></td>
<td style="text-align: center;"><p>10</p></td>
<td style="text-align: center;"><p>9</p></td>
<td style="text-align: center;"><p>8</p></td>
<td style="text-align: center;"><p>7</p></td>
<td style="text-align: center;"><p>6</p></td>
<td style="text-align: center;"><p>5</p></td>
<td style="text-align: center;"><p>4</p></td>
<td style="text-align: center;"><p>3</p></td>
<td style="text-align: center;"><p>2</p></td>
<td style="text-align: center;"><p>1</p></td>
</tr>
<tr>
<td style="text-align: center;"><p><strong>Y</strong></p></td>
<td style="text-align: center;"><p>0</p></td>
<td style="text-align: center;"><p>0</p></td>
<td style="text-align: center;"><p>0</p></td>
<td style="text-align: center;"><p>0</p></td>
<td style="text-align: center;"><p>1</p></td>
<td style="text-align: center;"><p>0</p></td>
<td style="text-align: center;"><p>1</p></td>
<td style="text-align: center;"><p>0</p></td>
<td style="text-align: center;"><p>1</p></td>
<td style="text-align: center;"><p>1</p></td>
</tr>
<tr>
<td style="text-align: center;"></td>
<td style="text-align: center;"><p><br />
<span class="emoji" data-emoji="raised_eyebrow">🤨</span></p></td>
<td style="text-align: center;"><p><br />
🤨</p></td>
<td style="text-align: center;"><p><br />
🤨</p></td>
<td style="text-align: center;"><p><br />
<span class="emoji" data-emoji="raised_eyebrow">🤨</span></p></td>
<td style="text-align: center;"><p><br />
<span class="emoji" data-emoji="nauseated_face">🤢</span></p></td>
<td style="text-align: center;"><p><br />
🤨</p></td>
<td style="text-align: center;"><p><br />
🤢</p></td>
<td style="text-align: center;"><p><br />
<span class="emoji" data-emoji="raised_eyebrow">🤨</span></p></td>
<td style="text-align: center;"><p><br />
<span class="emoji" data-emoji="nauseated_face">🤢</span></p></td>
<td style="text-align: center;"><p><br />
🤢</p></td>
</tr>
</tbody>
</table>

## **Discrete** Calibration **⚖️**

::::: columns
::: {.column width="50%"}
+:-----:+:----:+:----:+:----:+:----:+
| **p̂** | 0.11 | 0.15 | 0.18 | 0.29 |
+-------+------+------+------+------+
| **R** | 10   | 9    | 8    | 7    |
+-------+------+------+------+------+
| **Ŷ** | 0    | 0    | 0    | 0    |
+-------+------+------+------+------+
| **Y** | 0    | 0    | 0    | 0    |
+-------+------+------+------+------+
|       | \    | \    | \    | \    |
|       | 🤨   | 🤨   | 🤨   | 🤨   |
+-------+------+------+------+------+
:::

::: {.column width="50%"}
$$\begin{aligned}
\scriptsize{
\\\text{Observed: }\frac{\text{0}}{\text{4}} = 0}
\end{aligned}$$

$$\begin{aligned}
\scriptsize{
\\\text{Predicted: }\frac{\text{0.11 + 0.15 + 0.18 + 0.29}}{\text{4}} = 0.1825}
\end{aligned}$$
:::
:::::

## **Discrete** Calibration **⚖️**

::::: columns
::: {.column width="50%"}
+:-----:+:---:+:----:+:----:+
| **p̂** | 031 | 0.33 | 0.45 |
+-------+-----+------+------+
| **R** | 6   | 5    | 4    |
+-------+-----+------+------+
| **Y** | 1   | 0    | 1    |
+-------+-----+------+------+
|       | \   | \    | \    |
|       | 🤢  | 🤨   | 🤢   |
+-------+-----+------+------+
:::

::: {.column width="50%"}
$$\begin{aligned}
\scriptsize{
\\\text{Observed: }\frac{\text{2}}{\text{3}} = 0.66'}
\end{aligned}$$

$$\begin{aligned}
\scriptsize{
\\\text{Predicted: }\frac{\text{0.31 + 0.33 + 0.45}}{\text{3}} = 0.363'}
\end{aligned}$$
:::
:::::

## **Discrete** Calibration **⚖️**

::::: columns
::: {.column width="50%"}
+:-----:+:----:+:----:+:----:+
| **p̂** | 0.47 | 0.63 | 0.72 |
+-------+------+------+------+
| **R** | 3    | 2    | 1    |
+-------+------+------+------+
| **Y** | 0    | 1    | 1    |
+-------+------+------+------+
|       | \    | \    | \    |
|       | 🤨   | 🤢   | 🤢   |
+-------+------+------+------+
:::

::: {.column width="50%"}
$$\begin{aligned}
\scriptsize{
\\\text{Observed: }\frac{\text{2}}{\text{3}} = 0.66'}
\end{aligned}$$

$$\begin{aligned}
\scriptsize{
\\\text{Predicted: }\frac{\text{0.47 + 0.63 + 0.72}}{\text{3}} = 0.607}
\end{aligned}$$
:::
:::::

## **Discrete** Calibration **⚖️**

```{r}
probs <- c(0.3, 0.4, 0.4, 0.2, 0.1)
reals <- c(0, 1, 1, 1, 0)

plotly::plot_ly(
    heigt = 450,
    width = 450,
    x = ~Predicted, 
    y = ~Observed) |>  
  plotly::add_lines(
    name = "Perfect Calibration",
    data = data.frame(
      Observed = c(0, 1),
      Predicted = c(0, 1)),
      color = I("grey"), 
    line = list(dash="dot"),
    showlegend = FALSE
    ) |> 
  plotly::add_trace(
    data = data.frame(
      Observed = c(0, 2/3, 2/3),
      Predicted = c(0.1825, 0.363, 0.607)),
      color = I("black"), 
    showlegend = FALSE,
    name = "Discrete",
    mode = 'lines+markers',
    hovertemplate = paste0("<b>%{xaxis.title.text}:</b> %{x:.2f}<br>", 
                           "<b>%{yaxis.title.text}:</b> %{y:.2f}")) |> 
  plotly::layout(
    xaxis = list(showgrid = FALSE),
    yaxis = list(showgrid = FALSE),
    plot_bgcolor  = "rgba(0, 0, 0, 0)",
    paper_bgcolor = "rgba(0, 0, 0, 0)",
    fig_bgcolor   = "rgba(0, 0, 0, 0)"
    ) |> 
  plotly::config(displayModeBar = FALSE)
```

## **Discrete** Calibration **⚖️**

```{r}

library(rtichoke)

create_calibration_curve(
  size = 500,
  probs = list(example_dat$estimated_probabilities),
  reals = list(example_dat$outcome), type = "discrete"
)|> 
  plotly::layout(
    plot_bgcolor  = "rgba(0, 0, 0, 0)",
    paper_bgcolor = "rgba(0, 0, 0, 0)",
    fig_bgcolor   = "rgba(0, 0, 0, 0)"
    )

```

## **Smooth** Calibration **⚖️** {.smaller}

::::: columns
::: {.column width="40%"}
A different approach is to use a smoothing algorithm, in {rtichoke} I use *gam* for large samples and *lowess* for small samples.
:::

::: {.column width="60%"}
```{r}
create_calibration_curve(
  size = 500,
  probs = list(example_dat$estimated_probabilities),
  reals = list(example_dat$outcome), type = "smooth"
) |> 
  plotly::layout(
    plot_bgcolor  = "rgba(0, 0, 0, 0)",
    paper_bgcolor = "rgba(0, 0, 0, 0)",
    fig_bgcolor   = "rgba(0, 0, 0, 0)"
    )
```
:::
:::::

## **Smooth** Calibration **⚖️** {.smaller}

::::: columns
::: {.column width="40%"}
If you use smooth calibration, take a moment to explore the ranges of the curve!

It might look bad if you don't zoom-in to the reasonable range of estimated probabilities. That's why many times you might see Histograms or Rug-plots under the Calibration Curve.
:::

::: {.column width="60%"}
```{r}

probs_small_probs <- c(rep(0.1, 10), 
                         rep(0.2, 10), 
                         rep(0.15, 10), 
                         rep(0.25, 10),
                       0.7) + runif(min = -0.02,
                                                max = 0.02,
                                                n = 41)
reals_small_probs <- c(c(1, rep(0, 9)), 
c(1, rep(0, 9)), 
c(rep(1, 2), rep(0, 8)), 
c(rep(1, 2), rep(0, 8)), 0 )


create_calibration_curve(
  size = 500,
  probs = list(probs_small_probs),
  reals = list(reals_small_probs),
  type = "smooth"
)|> 
  plotly::layout(
    plot_bgcolor  = "rgba(0, 0, 0, 0)",
    paper_bgcolor = "rgba(0, 0, 0, 0)",
    fig_bgcolor   = "rgba(0, 0, 0, 0)"
    )

```
:::
:::::
